\documentclass[10pt,twocolumn]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{cite}

\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  breaklines=true,
  frame=single
}

\title{\textbf{CRYS-L: A Deterministic Execution Engine\\for Safety-Critical Engineering Computations}\\
{\large IEEE-754 Exact, Zero-Variance, Publicly Verifiable}}

\author{
  Percy Rojas Masgo \\
  Condesi Per\'{u} / Qomni AI Lab \\
  \texttt{percy.rojas@condesi.pe}
}

\date{April 2026}

\begin{document}

\maketitle

\begin{abstract}
Engineering certification standards---including NFPA~20, IEC~60364, and ASCE~7---require that
computational results be \emph{mathematically reproducible}: given identical inputs, every server
must produce bit-identical outputs under any load. This property, which we call
\emph{zero-variance determinism}, is incompatible with stochastic inference (LLMs), version-drifted
runtimes (Python/NumPy), and C++ code compiled without strict IEEE-754 flags.

We present \textbf{CRYS-L}, an execution engine that enforces zero-variance determinism through
three mechanisms: (1)~the \emph{branchless oracle pattern}, which eliminates undefined behavior
and enables SIMD vectorization of physical formulas; (2)~Cranelift JIT compilation targeting AVX2;
and (3)~strict IEEE-754 enforcement at the compiler level.
CRYS-L exposes 56 physics plans across 6 engineering domains via a public REST API. We report
6{,}515$\times$ throughput over Python/NumPy and 1.53$\times10^9$ scenarios/second on
commodity hardware (\$80/month, AMD EPYC 12-core). All results are publicly verifiable
without credentials at \url{https://desarrollador.xyz}.
\end{abstract}

%------------------------------------------------------------
\section{Introduction}
\label{sec:intro}

Safety-critical engineering calculations---fire protection sizing (NFPA~20), cable ampacity
(IEC~60364), structural load analysis (ASCE~7)---carry a requirement that is rarely discussed in
the systems literature: \emph{certified reproducibility}. An auditor must be able to re-execute
any calculation months later on any server and obtain the same IEEE-754 bit pattern.

This requirement immediately disqualifies three classes of tools widely used in practice:

\begin{enumerate}
\item \textbf{LLMs}: stochastic sampling, non-deterministic even at temperature 0 across different
  hardware~\cite{narayanan2023reproducibility}.
\item \textbf{Python/NumPy}: floating-point results change across NumPy versions, BLAS
  backends, and compiler optimization levels~\cite{goldberg1991floating}.
\item \textbf{Unsafe C++}: \texttt{if (flow < 1.0) return NAN} introduces a branch that
  (a)~prevents SIMD vectorization and (b)~propagates NaN silently through downstream
  calculations, constituting undefined behavior on invalid input.
\end{enumerate}

CRYS-L addresses this gap with a minimal design:\ physics expressions are
compiled to machine code via Cranelift~\cite{cranelift}, branches are eliminated via
predicate multiplication, and IEEE-754 is enforced as a first-class compiler invariant.

The system is live and verifiable. Every claim in this paper can be independently confirmed
with a single \texttt{curl} command requiring no account, no API key, and no installation.

%------------------------------------------------------------
\section{Background}
\label{sec:background}

\subsection{IEEE-754 and Determinism}

IEEE-754-2019 defines bit-exact results for the four arithmetic operations and square root,
but \emph{only when the same rounding mode, precision, and instruction sequence are used}.
Production JVM and Python runtimes frequently violate this at the library level through
auto-vectorization heuristics, fused multiply-add (FMA) contraction, and version-dependent
BLAS selection~\cite{monniaux2008pitfalls}.

CRYS-L enforces determinism through three compiler-level invariants:
(i)~all computations use IEEE-754 double precision (\texttt{f64});
(ii)~FMA contraction is disabled in the Cranelift backend;
(iii)~all control flow is eliminated before code generation (Section~\ref{sec:design}).

\subsection{Certification Requirements}

NFPA~20 (fire pump sizing), IEC~60364-5-52 (cable sizing), and ASCE~7-22
(structural loads) share a common audit requirement: given the same parameters,
the same result must be reproduced on demand. Existing software tools achieve this by
fixing versions and platform~\cite{nfpa20}. CRYS-L achieves it by construction,
independently of deployment environment.

%------------------------------------------------------------
\section{Design}
\label{sec:design}

\subsection{The Branchless Oracle Pattern}

The central design insight is that conditional validation can be expressed as a
floating-point predicate, eliminating branches entirely.

\textbf{Standard C++ formulation} (branched, UB on invalid input):
\begin{lstlisting}[language=C++]
float nfpa20_pump_hp(
    float flow, float head, float eff)
{
  if (flow < 1.0 || eff < 0.10)
    return NAN; // branch kills vectorization
  return (flow * 0.06309 * head * 0.70307)
         / (eff * 76.04);
}
\end{lstlisting}

\textbf{CRYS-L oracle formulation} (branchless, no UB):
\begin{lstlisting}
oracle nfpa20_pump_hp(
    flow: float, head: float, eff: float)
    -> float:
  let valid =
    (flow >= 1.0) * (flow <= 50000.0)
    * (eff >= 0.10)
  ((flow * 0.06309 * head * 0.70307)
   / (eff * 76.04 + 0.0001)) * valid
\end{lstlisting}

The predicate \texttt{valid} evaluates to \texttt{1.0} when all constraints hold and
\texttt{0.0} otherwise. The multiplication \texttt{result * valid} short-circuits to
\texttt{0.0} on invalid input without a branch, and the small epsilon \texttt{0.0001}
in the denominator prevents division-by-zero. No NaN is ever produced.

\textbf{Vectorization.} The branchless form allows AVX2 to evaluate 4 scenarios
in a single \texttt{VMULPD} instruction. The branched form prevents this optimization
because conditional branches inhibit SIMD vectorization in all major compilers.

\subsection{Cranelift JIT Compilation}

CRYS-L parses oracle definitions into a typed IR and emits Cranelift~\cite{cranelift}
intermediate representation directly. Cranelift provides:

\begin{itemize}
\item \textbf{AVX2 code generation} with explicit vector intrinsic selection.
\item \textbf{No floating-point contraction}: FMA is never emitted unless explicitly requested.
\item \textbf{Portable determinism}: the same Cranelift IR produces identical machine code
  across Linux kernel versions and CPU microcode variants (for the same ISA).
\end{itemize}

The JIT compilation pipeline is:
$$\text{CRYS-L source} \rightarrow \text{AST} \rightarrow \text{Cranelift IR} \rightarrow \text{AVX2 binary}$$

Compiled plans are cached in an oracle cache keyed by expression hash; subsequent
invocations incur zero compilation overhead.

\subsection{Pareto Optimizer}

CRYS-L includes a multi-objective Pareto sweep that evaluates the full parameter
space and returns all Pareto-optimal solutions. For a plan with $n$ parameters
each discretized to $k$ values, this explores $k^n$ scenarios in $< 2$\,ms
using the vectorized oracle on a single thread. On the NFPA~20 pump sizing plan
with flow $\in [361, 535]$\,gpm and head $\in [10, 150]$\,psi,
the sweep returns 127 Pareto-optimal configurations across efficiency and cost axes.

%------------------------------------------------------------
\section{Implementation}
\label{sec:impl}

\subsection{Server Architecture}

CRYS-L is implemented in 2{,}843 lines of Rust (reduced from an initial 15{,}122 lines
via modularization). The server exposes a raw HTTP/1.1 interface on port 9001 with no
framework dependency. All routing is performed by a single match expression on
\texttt{(method, path)} tuples.

The deployment runs on a \$80/month Contabo Cloud VPS (AMD EPYC 12-core, 48\,GB RAM,
500\,GB NVMe). Nginx proxies TLS termination and applies rate limiting for the
public demo tier.

\subsection{Determinism Proof Endpoint}

The \texttt{/verify} endpoint provides a live, independently verifiable proof of
zero-variance determinism:

\begin{lstlisting}[language=bash]
curl "https://desarrollador.xyz/verify?runs=20"
\end{lstlisting}

The endpoint executes the voltage-drop oracle 20 times and reports:
\begin{itemize}
\item \texttt{variance}: always \texttt{0.000000000000} (not approximately zero---exactly zero).
\item \texttt{result\_hash}: FNV-1a hash of the IEEE-754 \texttt{f64} bit pattern, identical across all runs and all invocations.
\item \texttt{all\_identical}: \texttt{true}.
\end{itemize}

\subsection{Plan Coverage}

Version 3.2 ships 56 physics plans across 6 domains:

\begin{table}[h]
\centering
\small
\begin{tabular}{lll}
\toprule
Domain & Standard & Plans \\
\midrule
Fire protection & NFPA 20 & 8 \\
Electrical & NEC / IEC 60364 & 12 \\
Structural & ASCE 7 & 9 \\
HVAC & ASHRAE & 11 \\
Finance & --- & 8 \\
Medical & IEC 60601 & 8 \\
\bottomrule
\end{tabular}
\caption{CRYS-L plan coverage by domain.}
\label{tab:plans}
\end{table}

%------------------------------------------------------------
\section{Evaluation}
\label{sec:eval}

All experiments are reproducible by any reader via the public API at
\url{https://desarrollador.xyz}. No registration or API key is required.

\subsection{Determinism Under Adversarial Input}

\begin{lstlisting}[language=bash]
curl https://desarrollador.xyz/simulation/adversarial
\end{lstlisting}

This endpoint executes 100{,}000 adversarial inputs (edge cases: zero flow, negative
efficiency, overflow values, subnormal floats) and reports:

\begin{itemize}
\item \textbf{Panics}: 0
\item \textbf{NaN propagated}: false
\item \textbf{Inf produced}: false
\item \textbf{Zero-division}: 0 (epsilon guard)
\end{itemize}

The branchless oracle pattern guarantees these properties by construction: invalid inputs
map to \texttt{0.0} through predicate multiplication, never to NaN or Inf.

\subsection{Jitter and Latency}

\begin{lstlisting}[language=bash]
curl https://desarrollador.xyz/simulation/jitter_bench
\end{lstlisting}

Across $10^4$ executions, we measure:

\begin{table}[h]
\centering
\small
\begin{tabular}{lr}
\toprule
Metric & Value \\
\midrule
Compute latency ($\sigma$) & 2{,}062\,ns \\
vs.\ C++ baseline (5\,M/s) & 850$\times$ \\
AVX2 utilization & 17.29\% \\
p99 latency & $< 10\,\mu$s \\
\bottomrule
\end{tabular}
\caption{Jitter benchmark results (live, AMD EPYC 12-core).}
\label{tab:jitter}
\end{table}

\subsection{Throughput vs. Alternatives}

\begin{lstlisting}[language=bash]
curl https://desarrollador.xyz/benchmark/vs_llm
\end{lstlisting}

\begin{table}[h]
\centering
\small
\begin{tabular}{lrrl}
\toprule
System & Throughput & Jitter $\sigma$ & Determinism \\
\midrule
\textbf{CRYS-L v3.2} & \textbf{1.53\,B/s} & \textbf{2{,}062\,ns} & IEEE-754 exact \\
C++ GCC -O3 & $\sim$5\,M/s & $\sim$850\,$\mu$s & UB on NaN \\
Python/NumPy & $\sim$0.2\,M/s & $>1$\,ms & float drift \\
GPT-4 Turbo & 0.08\,ans/s & $>1$\,s & stochastic \\
\bottomrule
\end{tabular}
\caption{Throughput comparison. CRYS-L runs 56 plans; C++ and Python run equivalent
single-formula implementations. GPT-4 measured via official API.}
\label{tab:throughput}
\end{table}

CRYS-L achieves 6{,}515$\times$ the throughput of Python/NumPy and
$1.53\times10^9$ scenarios/second, equivalent to evaluating the entire
Pareto front of an NFPA~20 pump sizing problem in under 2\,ms.

\subsection{SIMD Density}

\begin{lstlisting}[language=bash]
curl https://desarrollador.xyz/simulation/status
\end{lstlisting}

The status endpoint reports AVX2 utilization (SIMD instructions as fraction of total)
and confirms IEEE-754 enforcement mode. Current production values:
\texttt{ieee754: true}, \texttt{kernel: cranelift-avx2}.

%------------------------------------------------------------
\section{Related Work}
\label{sec:related}

\textbf{Deterministic floating-point.} Monniaux~\cite{monniaux2008pitfalls} surveys
pitfalls in floating-point reproducibility across compilers. Demmel and
Nguyen~\cite{demmel2013fast} propose reproducible floating-point summation.
CRYS-L extends these results to a full JIT execution pipeline for
domain-specific physics formulas.

\textbf{Domain-specific languages for physics.}
Liszt~\cite{devito2011liszt} and Regent~\cite{slaughter2015regent} provide DSLs for
mesh-based physics simulation with performance portability. CRYS-L targets a
complementary class---scalar closed-form engineering formulas---where
certification reproducibility is the primary constraint.

\textbf{JIT compilation for numerical code.}
LLVM~\cite{llvm} and Cranelift~\cite{cranelift} are used for numerical JIT.
Unlike LLVM, Cranelift does not perform FMA contraction by default, making it
preferable for IEEE-754 reproducibility.

\textbf{Approximate computing.}
TurboQuant~\cite{turbo2026} (ICLR 2026) achieves near-optimal distortion bounds
for lossy vector quantization. This is the \emph{complementary} problem to CRYS-L:
TurboQuant minimizes \emph{controlled} distortion; CRYS-L eliminates distortion entirely.
Both are valid engineering choices depending on whether exact certification is required.

%------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}

CRYS-L demonstrates that zero-variance determinism and high-throughput computation
are not competing goals. By eliminating branches at the language level and targeting
AVX2 via Cranelift JIT, CRYS-L achieves $1.53\times10^9$ scenarios/second with
provably zero floating-point variance---a property verifiable in real time by any reader.

The live API at \url{https://desarrollador.xyz} provides four independent proofs:
jitter benchmark, SIMD density, adversarial resilience, and determinism hash.
Source code is available at \url{https://github.com/condesi/crysl}.

%------------------------------------------------------------
\bibliographystyle{plain}
\bibliography{refs}

\end{document}
